Lexical frequency mediates the facilitation effect of cognateness in bilingual lexical acquisition

### 2023-04-17 manuscript.qmd
Fix comments and use apaquarto extension

---
title: "The role of cross-linguistic lexical similarity on bilingual word acquisition"
editor: source
author:
    - name: Gonzalo Garcia-Castro
      corresponding: true
      affiliations:
        - name: Center for Brain and Cognition, Universitat Pompeu Fabra
    - name: Daniela S. Ávila-Varela
      affiliations:
        - name: Center for Brain and Cognition, Universitat Pompeu Fabra
    - name: Ignacio Castillejo
      affiliations:
        - name: Departamento de Psicología, Universidad Autónoma de Madrid
    - name: Núria Sebastian-Galles
      affiliations:
        - name: Center for Brain and Cognition, Universitat Pompeu Fabra

pagetitle: Cross-linguistic similarity and  word acquisition
abstract: |
    Bilinguals face the challenging task of learning words from languages with overlapping phonologies. Floccia et al. (2018) reported larger vocabulary sizes for 24-month-old bilinguals that were learning languages that shared a greater amount of cognates (e.g., English-Dutch). The mechanisms underlying this effect remain unknown. We explore two compatible scenarios. First, we test whether cognates are learnt earlier than non-cognates. This would account for the difference in vocabulary size associated to the amount of shared cognates across languages. Second, we explore the possibility that the word-forms of one language interact with those form the other language, scaffolding the acquisition of their translation equivalents when their phonologies overlap. This mechanism, in line with the parallel activation account of bilingual speech perception, would provide a plausible explanation to why cognates are acquired ealier by bilinguals. We developed an online tool to collect parental reports of receptive and productive vocabularies from children learning Catalan and/or Spanish, and present data on receptive and productive vocabulary of bilingual toddlers aged 10 to 36 months.
thanks: |
    This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019). The authors declare no conflicts of interest with regard to the funding source of this study. We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We would like to thank the clinics Quirón and Sagrada Familia that allowed us to recruit participants in their premises. We also thank all families and infants who participated in the experiments. Data collection was half-way when the COVID-19 pandemic started. We would like to pay special tribute to the families that collaborated with us under these difficult circumstances.
keywords: lexical acquisition, vocabulary, bilingualism
toc: false
number-sections: true
fig-dpi: 1000
csl: apa.csl
bibliography: manuscript/references.bib
google-scholar: true
format:
  pdf:
    pdf-engine: xelatex
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    keep-tex: true
    linestretch: 1.25
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
    echo = FALSE,
    fig.align = "center",
    message = FALSE,
    warning = FALSE,
    cache.extra = knitr::rand_seed,
    out.width = "100%",
    dpi = 800,
    results = "asis"
)
```

```{r}
#| label: packages
#| echo: false
#| message: false
#| warning: false
#| include: false
# load packages
library(brms) 
library(conflicted)
library(dplyr) 
library(forcats)
library(ggplot2) 
library(ggsci)
library(gt) 
library(here) 
library(janitor)
library(knitr)
library(lubridate)
library(multilex)
library(patchwork) 
library(papaja) 
library(purrr)
library(scales) 
library(stringr)
library(tidybayes) 
library(tidyr) 
library(tibble)

# load and process objects
tar_load(participants)
tar_load(items)
tar_load(multilex_data)
tar_load(model_log_liks)
tar_load(model_loos)
tar_load(model_fit_1)
tar_load(rope_coefs)
tar_load(rope_interval)
tar_load(df)
tar_load(posterior_description)

participants <- participants %>% 
    left_join(select(multilex_data$logs, id, time, edu_parent))
items <- items %>% 
    left_join(select(pool, item, language, category, class))
# set global options
options(
    knitr.kable.NA = "-",
    knitr.duplicate.label = "allow",
    gt.html_tag_check = FALSE,
    ggplot2.discrete.fill = ggsci::pal_d3()(4),
    ggplot2.discrete.colour = ggsci::pal_d3()(4),
    ggplot2.continuous.fill = ggplot2::scale_color_gradient,
    ggplot2.continuous.colour = ggplot2::scale_color_gradient
)

# set custom ggplot2 theme
theme_set(theme_custom())

# resolve namespace conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
```

# Introduction {#introduction}

The foundations of word learning are in place early in age: before the end of their first year of life infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999some; @bergelson2012months; @bergelson2015early] and parental reports [e.g., @fenson2007macarthur]. During the last half of their second year, they acquire new words at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002children; @bergelson2020comprehension]. These early stages of lexical acquisition are characterised by substantial variation across children, reflected, for instance, on the variation on the number of words they know [i.e. vocabulary size, @fenson1994variability; @frank2021variability] or on the proportion of those words that fall into the category of nouns, as opposed to verbs, adjectives, or function words [e.g., @nelson1973structure, bates1994developmental].

Despite this variability, children's trajectories of vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10 to 16 month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caretakers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animals (*woof-woof*). These results were later extended by @frank2021variability to a wider diversity languages, also reporting that such cross-language commonalities are stronger at earlier stages of lexical acquisition, compared to later stages. Most of the literature on early word acquisition, however, has been conducted on monolingual children, and neglects the problem of how bilinguals--who represent a substantial proportion of the population in most societies--acquire words at early ages.

There is evidence that bilinguals know, on average, less words in each of their languages than monolinguals, but also that both groups know a similar amount of words when the two languages are aggregated. For example, @hoff2012dual found that English-Spanish bilingual toddlers in South Florida knew less words in English than monolinguals, who only learnt English. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, pointing to the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar--or even larger--number of words than monolinguals, only when the languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez2020bilingual]. While these studies have mostly relied on samples of bilingual children learning two relatively distant languages, as it is the case of English and Spanish, it is unclear whether children learning typologically more similar languages also know less words in each of their languages than monolinguals. What role could linguistic distance play during early vocabulary growth?

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms--one in each language. Depending on the linguistic distance between both languages, the two sets of words may overlap in varying degrees. Particularly, when both languages are typologically close, like Spanish and Catalan (both Roman languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two linguistically distant languages, like Spanish and English (one Roman, the other Germanic). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might hear *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning a pair of languages that share a high proportion of cognates should benefit more often from this facilitation effect than those learning a pair of languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence in line with this claim. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one a pool of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological similarity between the words in each of these additional languages and their translation equivalents in English. Phonological similarity was measured by computing the Levenshtein distance between each cross-language pair of phonological transcriptions. The Levenshtein distance is a metric that computes the edit distance between two strings by counting the smallest number of insertions, deletions and substitutions one of the strings has to go through to become identical to the other [@levenshtein1966binary]. The resulting scores were then divided by the length of the longest string to bound the similarity scores between 0 and 1, and then entered this variable as a predictor as they modelled participants' vocabulary sizes. Among other findings, the authors reported an increase in productive vocabulary size in the additional language associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (22.14% phonological similarity), were able to produce more Dutch words than English-Mandarin bilinguals (1.97% phonological similarity) were able to produce in Mandarin.

Floccia et al. pointed to *parallel activation* as the main mechanism underpinning their results. The parallel activation hypothesis suggests that bilinguals activate both languages simultaneously during speech production or comprehension, and that this phenomenon is the result of the activation of lexical representations in both languages, even when only one is in use during production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain]. One of the clearest examples of parallel activation was provided by @costa2000cognate. In this study, Catalan-Spanish monolingual and bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trials labels were non-cognates (*mesa*-*taula*, translations of *table*). Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. Importantly, Spanish monolinguals--unfamiliar with the Catalan translations of the Spanish words they uttered--did not show this effect. These results suggest that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Several subsequent studies have also provided similar evidence in comprehension in children [e.g., @von2012language; @poulin2013lexical]. Parallel activation is therefore a plausible mechanism to account for Floccia et al.'s results: cognates increase the amount of activation in both languages, facilitating word acquisition, and ultimately leading to children learning language pairs with a larger proportion of cognates are predicted to show larger vocabulary sizes. A missing piece in this account is how increase parallel activation might facilitate the acquisition new words.

In this study, we hypothesise that cognates are acquired earlier in age than non-cognates, which might explain why children learning languages that share many cognates show larger vocabulary sizes. Evidence supporting this claim is scarce. @bosch2014first used vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect reported a longitudinal single case of one English-German bilingual who produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. On the other hand, Floccia et al.'s estimates are statically more reliable given their (much larger) sample size, but their study was not aimed at testing the effect of cognateness on age of acquisition directly. In their discussion the authors state the following (pp. 70):

> "This  finding  also  provides  support  to  the  proposal  that  the  cognate advantage is due to cognates being acquired before non-cognates in early childhood (Costa et al., 2016), leading to an ease of processing later in life."

We identify two main reasons why an earlier age of acquisition for cognates than for non-cognates is an unwarranted conclusion from Floccia et al.'s results. First, the response variable used was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all items into a single datum per child, information about the acquisition status of cognates vs. non-cognates was lost. Second, all participants were aged ~24 months, meaning that even if the unaggregated responses to individual items were included as response variable, the possible effect of cognateness could only be interpreted as an increase or decrease in the likelihood of participants at such age to have acquired each item, and not as an increase or decrease in the age of acquisition of such item.

More recently, @mitchell2022cognates addressed this issue overcoming some of these pitfalls. Using a larger, longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size grows faster when both languages are phonologically more similar: word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster.

As @mitchell2022cognates note in their discussion, two competing hypotheses remain untested. The facilitative effect of cognateness can be explained by, at least, two alternative, arguably incompatible, mechanisms. On the one hand, cognateness might operate as a facilitator of the acquisition of both forms of a translation pair, with every exposure to either form activating the other, and therefore strengthening the association between both and their referent. If this is the case, then cognateness should increase the probability of acquisition of the forms in both languages equally.

Another possibility is that cognateness plays a role in the acquisition of a translation pair only when at least one of the word-forms has been acquired. This account departs from the premise that, if the child is not previously familiar with one of the word forms, the acquisition of its translation in the other language cannot be facilitated by cognateness. This implies that, if cognateness facilitates the acquisition of a word form, it can only do so after the child has acquired one of the word forms of the translation pair. Given that children are more likely to acquire words from languages to which they are exposed more often [@david2008individual; @cattani2014much; @thordardottir2006bilingual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness.

In line with this second hypothesis, the vocabulary size associated with linguistic similarity that Floccia et al. reported was larger in the additional language vocabulary than in English vocabulary. Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have, on average, learnt the English word-form of translation equivalents earlier than the word-form in the additional language.  If this is the case, then the acquisition of English words by English-dominant participants would rarely benefit from their cognate status (the other word-form is not available yet), while the acquisition of words in the additional language would benefit from their phonological similarity with the (available) English form.

In this study, we investigated the role of cognateness on the acquisition of translation equivalent, and whether such effect is conditional to the amount of time the child is exposed to the corresponding language of each member of the translation pair. Using an online vocabulary checklist--designed specifically for this study--we collected data from a sample of children aged `r min(floor(participants$age))` to `r max(floor(participants$age))` months learning Catalan and/or Spanish, with varying degrees of exposure to each language. We then adopted an Item Response Theory (*IRT*) approach to model the probability of participants being reported by their parents to understand or understand *and* say each word in the checklist, conditional to its cognate status in Catalan and Spanish and participants' degree of exposure to the corresponding language, while adjusting for other indices of word difficulty--lexical frequency and length (phonemes)-- and participant skill--age [see @kachergis2022toward for a similar approach].

# Methods {#methods}

Data and materials are available at OSF [https://osf.io/hy984/](https://osf.io/hy984/), and code is available at GitHub [https://github.com/gongcastro/trajectories](https://github.com/gongcastro/trajectories).

## Participants {#participants}

```{r}
#| label: participants
#| echo: false
#| message: false
#| warning: false
#| include: false
# load packages
n_participants <- length(unique(participants$id))
dates <- as_date(range(participants$time_stamp, na.rm = TRUE))
```

We collected data from `r printnum(n_participants)` children from the Metropolitan Area of Barcelona between `r format(dates[1], "%dth %B, %Y")` and `r format(dates[2], "%dth %B, %Y")`. All families gave informed consent before participating and this study was approved by the Comitè d'Ètica de la Investigació amb Medicaments (CEIm) from Hospital del Mar (Barcelona, Spain), code XXXXXXXXX. Families from the participant database of the Laboratori de Recerca en Infància of the Universitat Pompeu Fabra were contacted by e-mail or phone if their child were aged between `r floor(min(participants$age))` and `r floor(max(participants$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device in a browser within the two week following the invitation to participate. @tbl-participants-lp summarises the distribution of participants across ages and degrees of exposure (DoE) to Catalan.

>>>>

---
title: "Lexical frequency mediates the facilitation effect of cognateness in bilingual lexical acquisition"

shorttitle: Frequency and cognateness in lexical acquisition

author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias Fargas 25-27
        city: Barcelona
        region: Spain
        postal-code: "08005"

  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Facultade de Letras
        address: Alameda da Universidade
        city: Lisboa
        region: Portugal
        postal-code: "1600-214"

  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        address: Iván Pavlov, 6
        city: Madrid
        region: Spain
        postal-code: "28049"

  - name: Núria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations: 
      - ref: cbc

csl: "assets/apa7.csl"
bibliography: "assets/references.bib"

editor: source

format:
  apaquarto-docx: default
  apaquarto-pdf:
    pdf-engine: xelatex
    documentmode: man
    fontsize: 12pt
    keep-tex: false
    floatsintext: true
    blank-lines-above-title: 2
    header-includes:
      - \usepackage{tipa}
      - \usepackage[pinyin]{babel}
      - \usepackage{CJKutf8}
      - \usepackage{fontspec}
      - \babelprovide[script=CJK, language={Chinese Simplified}]{chinese-simplified}
      - \babelfont[chinese-simplified]{rm}[Ligatures = Common]{Noto Serif CJK SC}
      - \babelfont[chinese-simplified]{sf}[Ligatures = Common]{Noto Sans CJK SC}

    include-in-header:
      - text: '\usepackage{lineno}\linenumbers'

author-note:
  blank-lines-above-author-note: 3
  disclosures:
    conflict-of-interest: |
        The authors declare no conflicts of interest with regard to the funding source of this study.
    financial-support: |
        This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019).
    gratitude: |
        We are grateful to Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, Ege E. Özer, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support. We also thank Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.

abstract: |
    Language non-selectivity is prominent feature of bilingual lexical processing. An instance of such non-selectivity is embodied by cognateness, which recent studies have suggested to facilitate vocabulary acquisition in bilingual toddlers, who show larger vocabulary sizes when their languages share many cognates, and to be acquired at earlier ages than non-cognates. The specific mechanisms behind such facilitation are unclear. We present an account of early bilingual vocabulary growth in which cognateness interacts with lexical frequency and language exposure to facilitate the word acquisition. We evaluated this model against vocabulary data from 436 Catalan-Spanish bilinguals aged 12 to 34 months. We used Bayesian Exploratory Item Response Models to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the toddler. We found converging evidence for an earlier age-of-acquisition for cognate words, and for such effect being mediated by lexical frequency and language exposure. Low-frequency words, and words from the language of least exposure were more strongly benefitted by their cognate status than high-frequency words. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition of words in the other language.

keywords: [lexical acquisition, vocabulary, bilingualism, item response theory, bayesian]

fig-dpi: 1000
---
{{< include _extensions/wjschne/apaquarto/_apa_title.qmd >}}

```{r setup}
#| label: setup
#| echo: false
#| message: false
#| warning: false
# load packages ----------------------------------------------------------------
tar_load_globals()

# load objects -----------------------------------------------------------------
tar_load_everything()

items <- bvq_data$pool  |>  
    mutate(item = str_remove(item, "cat_|spa_")) |> 
    select(item, language, semantic_category, class) |> 
    right_join(items, by = c("item", "language"))

# set ggplot theme and colour palette ------------------------------------------

theme_set(theme_custom())

bayesplot_theme_set(theme_custom()  +
                        theme(panel.grid = element_blank(),
                              axis.text = element_text(size = 5),
                              strip.text = element_text(size = 5),
                              plot.title = element_text(size = 8),
                              plot.subtitle = element_text(size = 8)))

clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")

# set params -------------------------------------------------------------------

color_scheme_set(rep(clrs[1], 6))
options(ggplot2.ordinal.fill = clrs,
        ggplot2.ordinal.colour = clrs,
        ggplot2.discrete.fill = clrs,
        ggplot2.discrete.colour = clrs,
        ggplot2.continuous.fill = scale_color_gradient,
        ggplot2.continuous.colour = scale_color_gradient,
        knitr.graphics.error = FALSE) 
```

# Introduction

The foundations of word learning are in place early in age. From six months of age, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [@fenson2007macarthurbates; @samuelson2021precision]. During the last half of their second year, new word forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. Such early stages of lexical acquisition are characterised by substantial variation across children reflected, for instance, on the variability of the number of words they know [i.e., vocabulary size, @fenson1994variability; @frank2021variability]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10-- to 16--month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caregivers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animal sounds (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to substantial proportion of the world population acquiring more than one language since early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical or the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. Depending on the similarity between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might be exposed to the words *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence in line with this claim. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological overlap was measured by computing the Levenshtein distance between each cross-language pair of phonetic transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters/phonemes by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/kæt/-/\textipa{kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /æ/ by /\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /kæt/-/\textipa{Gata}/) would score three, as it requires at least three editions: substituting /k/ for /\textipa{G}/, substituting /æ/ for /a/ and adding another /a/.

Levenshtein distance scores are divided by the length of the longest string to normalise the similarity scores between 0 and 1. Finally, the result is subtracted from one, to get the complementary, and therefore a measure of phonological similarity, instad of dissimilarity. Following the previous examples: $1/4 = 0.25$ for /kæt/-/\textipa{kAt}/, and $3/4 = 0.75$ for /kæt/-/\textipa{Gata}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (0.2214 mean phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (0.0197 mean phonological overlap) were able to produce in Mandarin.

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during speech production [@costa2000cognate; @hoshino2008cognate] or comprehension [@spivey1999cross; @thierry2007brain], even in fully monolingual environments. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), importantly such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results showed that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Regarding word acquisition, the parallel activation account is a plausible explanation for Floccia et al.'s results in the light of two observations. First, bilingual toddlers' lexicon seems to consist of a large proportion of translation equivalents: the presence of a word in their lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage]. Second, there is experimental evidence of parallel activation during word comprehension in children and toddlers [@vonholzen2012language; @poulin-dubois2013lexical; @bosma2020cognate]. It is possible that cognateness increases the amount of cross-language activation, facilitating word acquisition. This could ultimately lead to children who are learning languages with a larger proportion of cognates (e.g., English and Dutch) showing larger vocabulary sizes. This account in which cognateness underlies the facilitation effect of typological language distance is also in line with previous studies suggesting that the acquisition of new words is facilitated by their phonological or semantic similarity with other words [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children]. If phonological similarity plays a cross-language facilitation role during word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity (e.g., are taxonomically related) but not phonological similarity.

Evidence supporting an earlier age of acquisition for cognates is scarce. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. Floccia et al.'s estimates are statistically more reliable given their remarkably larger sample size (372 participants). Although their results suggest a cognateness effect on vocabulary size, their study was not aimed at testing the effect of cognateness on age of acquisition at the *word-level*. The response variable in this study was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all words into a single datum per child, it was no longer possible to test the effect of cognateness on word acquisition at the word-form level. Also, all participants were aged 24 months, meaning that even if the unaggregated responses to individual items were included as the response variable, age-sensitive claims about the effect of cognateness on word acquisition should be taken with caution [a limitation also present in the @bosch2014first study].

More recently, @mitchell2022cognates addressed this issue providing a more fine grained measure of a  cognateness facilitation effect in vocabulary acquisition. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age of acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size seems to grow faster when both languages are more similar at the lexical level. Word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

In the present investigation we approach the cognate facilitation on word learning by extending the conceptual framework of accumulator models of language acquisition to the bilingual case. Accumulator models formalise the widespread assumption that a child's vocabulary growth rate is a function of mainly---but not exclusively---the number of learning instances that a child has accrued with a particular word [see @kachergis2022standard for review]. A learning instance of a word consists, for example, in hearing such word-form (*flower*) in the presence of its referent (a flower). As infants receive linguistic input, they also accumulate opportunities to learn words until a threshold is crossed, an the word can be considered to be acquired. In line with this account, the frequency with which a child encounters a particular word-form in their speech input is a critical determiner of the age at which such word will be acquired. The more frequently [@hidaka2013computational; @mollica2017how]. For instance, words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are more likely to be acquired earlier than words with low lexical frequency such as 'taxi' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words) [e.g., @goodman2008does]: infants might accumulate learning experiences with 'flower' faster than with 'taxi'.

Previous accounts on vocabulary growth based on accumulator models are a convenient framework to study bilingual vocabulary growth, as bilinguals share fundamentally similar learning mechanisms with monolinguals. Developmental divergences between both groups can be understood as the result of the the negotiation between such mechanisms and the particular demands of a dual language input [e.g., @werker2005primir; @curtin2011bilingual]. One of such divergences is the fact that bilinguals' frequency of exposure to a word-form is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual would (who listens 100% of the time to English).

We suggest that the cognateness effect proposed by Floccia et al. and Mitchell et al. facilitates word acquisition by reducing the number of learning instances needed by a word to be acquired, and that this effect takes place *only* once the (cognate) translation of such word has been acquired. We hypothesise that both forms in the translation equivalent (e.g., *gat* and *gato* [Catalan and Spanish for 'cat'], for a Catalan-Spanish bilingual) follow their acquisition trajectory as they would for a monolingual child: accumulating learning instances, which occur a a rate proportional to the lexical frequency of the word-form and the amount of exposure to the language they belong to. When one of the word-forms is sufficiently acquired, the number of learning instances left needed for the acquisition of the other word-form is reduced.
